---
title: "Netflix Movies and TV Shows"
description: |
  Looking at Netflix content types from TidyTuesday
author: Eli Della Bitta
date: September 2, 2025
format: 
  html:
    warning: false
    message: false
    code-fold: true
---

## Introduction

This dataset contains information about movies and TV shows available on Netflix. I will create a simple visualization to see how much content Netflix has for movies versus TV shows.

## Data Loading

```{r}
#| label: setup
library(ggplot2)
library(dplyr)

netflix <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-04-20/netflix_titles.csv')
```

```{r}

content_counts <- netflix |>
  count(type)


ggplot(content_counts, aes(x = type, y = n, fill = type)) +
  geom_col() +
  labs(title = "Netflix Content: Movies vs TV Shows",
       x = "Content Type",
       y = "Number of Titles") +
  theme_minimal() +
  scale_fill_manual(values = c("Movie" = "red", "TV Show" = "black"))
```

## Data Sources

-   Tinytuesday Dataset: [Week 17, 2021 - Netflix Titles](https://github.com/rfordatascience/tidytuesday/blob/main/data/2021/2021-04-20/readme.md)
-   Original Data Source: Kaggle dataset "Netflix Movies and TV Shows" compiled by Shivam Bansal, originally sourced from Flixable (a third-party Netflix search engine)
